Composable data visualizations

### 2020-05-02 wip-compost/paper.tex
Nesting of scales, abstractions code

\title{Composing data visualizations}
\author[Tomas Petricek]{TOMAS PETRICEK\\
       University of Kent, UK\\
       \email{t.petricek@kent.ac.uk}}

@@

\begin{figure}
\begin{equation*}
\begin{array}{rcl}
l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
\end{array}
\end{equation*}
\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
\label{fig:scale}
\end{figure}

@@

\begin{figure}
  \begin{equation*}
  \begin{array}{rclcl}
  s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
    & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
  \end{array}
  \end{equation*}
  \caption{Additional combinators for controlling scales; \kvd{roundScale} and \kvd{explicitScale}
    can be used for creating nice scale ranges; \kvd{nest} makes it possible to create charts with
    multiple scales.}
\label{fig:control}
\end{figure}

The two combinators for controlling the range of scales are \kvd{roundScale} and \kvd{explicitScale}.
The oprations can be applied to either the X scale or the Y scale, which is indicated by the
$x/y$ sub-script. The \kvd{roundScale} primitve takes the inferred X or Y scale of the shape $s$
and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
For example, if a continuous scale has minimum $\num{0}$ and maximum $\num{365}$, the resulting
scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
The \kvd{explicitScale} operation is similar, but it replaces the inferred scale with an explicitly
provided scale (the type of the inferred scale has to match with the type of the explicitly given
scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
following code (reusing the \ident{labour} and \ident{conservative} variables defined earlier):
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
\qquad \kvd{overlay}~[~\ident{labour}, \ident{conservative}~]~)))\\
\end{array}
\end{equation*}

@@

@@

@@

codezzz
%
\begin{equation*}
\begin{array}{l}
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{continuous}~\num{0},\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{continuous}~\num{50},\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

moar codezzzz

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{brexitRate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\\
\quad \kvd{overlay}~[\\
\qquad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
\qquad \kvd{fill}~\strf{\#D6272860}, [\;\,\kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}\;],\\
\qquad \kvd{line}~\strf{\#202020}~\ident{rates}~]\\[0.5em]
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{continuous}~\num{0},\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{continuous}~\num{50},\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

\begin{figure}
  \includegraphics[scale=0.57]{figures/lines}
  \vspace{0.25em}
  \caption{TBA.}
  \label{fig:lines}
\end{figure}

@@

\begin{figure}[!h]
\includegraphics[scale=1,trim={0cm 7.5cm 6cm 0cm},clip]{nest.pdf}
\caption{A continuous scale with values from $0$ to $6$, nested in another scale.}
\label{fig:nesting}
\end{figure}

@@

Finally, one more primitive that we added in the above definition and that we have not explained
yet is $\kvd{nest}_{x/y}$. The parameters of the primitive are two values, $v_{min}$ and $v_{max}$
together with a shape $s$. The primitive makes it possible to nest one scale inside another
one.
When inferring the scales, Compost infers the scale of the nested shape $s$. This will then take
the area determined by the two points $v_{min}$ and $v_{max}$ in the scale determined by other
shapes that are overlaid with the $\kvd{nest}$ shape. The two points are also used to determine
the scales of the overall shape, but the scales of $s$ are nested and do not have any effect on
the outer scales. In fact, the scale of $s$ does not even have to be of the same type as the
scale of the outer shape.
Figure~\ref{fig:nesting} illustrates this with an example. Here, we are nesting a continuous scale
with values ranging from $\num{0}$ to $\num{6}$ into another scale. The values $v_{max}$ and
$v_{min}$ do not even have to be continuous. We can, for example, nest a line chart inside a
bar of a bar chart and then the two values could be $\kvd{cat}~\strf{One}, 0$ and $\kvd{cat}~\strf{One}, 1$
(which are two values that define a region of a categorical scale).
One area where nesting of scales is particularly useful is when combining multiple charts in
a single view. For example, let's say that we have two line charts showing prices of two different
stocks over the same period of time (for simplicity, we can just use UNIX timestamp as a time).
The prices of the two stocks are orders of magnitude different, so they cannot fit easily into the
same chart. We want to show two line charts side-by-side (one above each other). They should each
have their own Y axis (price), but the X axis (time) should be shared.
Assuming we have \ident{googPrices} and \ident{fbPrices} as two line charts without any axes,
covering the same date range, we can write:
\begin{equation*}
\begin{array}{l}
\kvd{axis}_x~(\kvd{overlay}\\
\quad (\kvd{nest}_y~(\kvd{cat}~\strf{top-chart}, \num{0}), (\kvd{cat}~\strf{top-chart}, \num{1}),
 (\kvd{axis}_y~\ident{fbPrices})),\\
\quad (\kvd{nest}_y~(\kvd{cat}~\strf{bottom-chart}, \num{0}), (\kvd{cat}~\strf{bottom-chart}, \num{1}),
 (\kvd{axis}_y~\ident{googPrices}))~)\\
\end{array}
\end{equation*}
Reading the code from the inner-most part to the outer-most, we first add separate Y axes to both
of the line charts. Given the difference in the prices, the axes will have quite different values.
We then nest the (continuous) Y axes using the \kvd{nest} primitive and, at the same time, implicitly
define a new outer Y axis. The outer Y axis is categorical with just two values, \strf{top-chart}
and \strf{bottom-chart}. This means that the two nested charts will take equal amount of space,
one above the other (each of the charts takes the full space allocated for the category -- the
minimal value has ratio $0$ while the maximal value has ratio $1$).
~
~
\noindent
[TODO: We really need illustrations for those example charts, but that would be more work!]
~
[TODO: Maybe use this chart as a motivation: https://wellcomeopenresearch.org/articles/4-63/v1]
~
\section{Random ideas}
\begin{itemize}
\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.
\item It would be worth thinking about ordinal values, which are categorical but can be sorted.
\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
 or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
 every value (including bar chart colors) should be coming from some scale...
\item Implementing something like \kvd{axis} or \kvd{roundScale} as an actual derived primitive is
 a bit tricky, because it needs to invoke a part of the normal rendering workflow (to run the
 automatic inference of scales on the nested shape) -- this might be just implementation issue,
 but it could be some more basic problem.
\end{itemize}
\end{document}

>>>>

\title{Composable data visualizations}
\author[Tomas Petricek]{TOMAS PETRICEK\\
       University of Kent, UK\\
       \email{t.petricek@kent.ac.uk}}

@@

\begin{figure}
\vspace{-0.5em}
\begin{equation*}
\begin{array}{rcl}
l & = & \kvd{continuous}~n_{min}, n_{max} \lsep \kvd{categorical}~[\,c_1, \ldots, c_k\,]
\end{array}
\end{equation*}
\vspace{-1em}
\caption{A scale $l$ can be continuous, defined by a range, or categorical, defined by a list of values.}
\label{fig:scale}
\vspace{-1em}
\end{figure}
\begin{figure}
\begin{equation*}
\begin{array}{rclcl}
s & = & \kvd{roundScale}_{x/y} s      &|& \kvd{nest}_{x/y}~v_{min}, v_{max}, s \\
  & | & \kvd{explicitScale}_{x/y}~l, s &|&  (\ldots)
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Additional combinators for controlling and nesting scales, extending earlier definition of $s$.}
\label{fig:control}
\end{figure}

@@

The two combinators for controlling the range of scales are \kvd{roundScale} and \kvd{explicitScale}.
The oprations can be applied to either the X scale or the Y scale, which is indicated by the
$x/y$ sub-script. The \kvd{roundScale} primitve takes the inferred X or Y scale of the shape $s$
and, if it is a continuous scale, rounds its minimal and maximal values to a ``nice'' number.
For example, if a continuous scale has minimum $\num{0}$ and maximum $\num{365}$, the resulting
scale would have a maximum $\num{400}$. For categorical scale, the operation does not have any effect.
The \kvd{explicitScale} operation is similar, but it replaces the inferred scale with an explicitly
provided scale (the type of the inferred scale has to match with the type of the explicitly given
scale). For example, the chart in Figure~\ref{fig:simple} (right) is constructed using the
following code (reusing the \ident{labour} and \ident{conservative} variables defined earlier):
%
\begin{equation*}
\begin{array}{l}
\kvd{axis}_l~(\kvd{axis}_b~(\kvd{roundScale}_y~(\kvd{explicitScale}_x~(\kvd{categorical}~[\,\strf{Labour},\strf{Conservative}\,]),\\
\qquad \kvd{overlay}~[~\ident{labour}, \ident{conservative}~]~)))\\
\end{array}
\end{equation*}
\vspace{-0.5em}

@@

\begin{figure}
\includegraphics[scale=0.57]{figures/lines}
\vspace{0.25em}
\caption{Two charts showing currency exchange rates with a shared X scale and separate Y scales.}
\label{fig:lines}
\end{figure}

@@

The most interesting primitive for controlling scale composition defined in Figure~\ref{fig:control}
is $\kvd{nest}_{x/y}$. The combinator takes two values, $v_{min}, v_{max}$ and a shape $s$ as arguments
and it nests the scale of the shape $s$ inside the region defined by $v_{min}, v_{max}$. When inferring
scales of shapes, the scale of $\kvd{nest}_{x/y}~l, s$ will be a categorical or continuous scale
inferred using the values $v_{min}$ and $v_{max}$, regardless of the values that are used inside
the shape $s$. The chart space between $v_{max}$ and $v_{min}$ will then be used to render the
nested shape $s$ using its inferred scale. An example of nesting is shown in Figure~\ref{fig:nesting}.
Here, a chart with a continous scale from $\num{1.1}$ to $\num{1.4}$ (e.g.~GBP-EUR exchange rates)
is nested in the left half of another chart, which has a continuous scale from $\num{0}$ to $\num{100}$.

\begin{figure}
\includegraphics[scale=1,trim={0cm 7.5cm 6cm 0cm},clip]{figures/nest}
\caption{A continuous scale with values from $0$ to $6$, nested in another scale.}
\label{fig:nesting}
\end{figure}

The nesting of scales can be used in a variety of ways. We can, for example, nest a line chart
inside a bar of a bar chart. In that case, the values for $v_{min}$ and $v_{max}$ would be
$\kvd{cat}~\strf{ABC}, 0$ and $\kvd{cat}~\strf{ABC}, 1$, which define the start and the end of
the region allocated to the $\strf{ABC}$ category on a categorical scale. A simpler use case for
the combinator is showing multiple charts in a single view. For example, the motivating example
in Figure~\ref{fig:charts} (right) comapres aligned line charts of exchange rates for two
different currencies. Assuming \ident{gbpusd} and \ident{gbpeur} are lists containing days as X
values and exchange rates as Y values, we can construct a simple chart with two line charts,
shown in Figure~\ref{fig:lines} (left), using:
%
\begin{equation*}
\begin{array}{l}
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\kvd{line}~\strf{\#202020}~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

\vspace{-0.5em}
\noindent
In this example, the X scale shows the days of the year. This scale is shared by both of the charts.
Indeed, if data was only available for the second half of the month for one of the charts,
we would want the line to start in the middle of the chart. However, the Y scale needs to be
separate for each of the charts. To achieve this, we use $\kvd{nest}_y$. The scale of the inner
shapes is continuous, from the minimal to the maximal exchange rate for a given period. The
outer scale is determined by the explicitly defined points. For the upper chart, these are
$\kvd{cont}~\num{0}$ and $\kvd{cont}~\num{50}$; for the lower chart, these are
$\kvd{cont}~\num{50}$ and $\kvd{cont}~\num{100}$. The continuous values define a scale that only
contain two shapes -- one in the upper half, one in the lower half -- and so the three numbers could
have equally been, for example, $\num{0},\num{1}, \num{2}$. The outer scale used here is
synthetic and it is not aligned with other chart elements. An example of a more complex chart
that follows a similar style, but does not have synthetic outer scale would be pairplot from
the seaborn Python library. % \cite

For completeness, the following code snippet shows how to construct the full currency exchange
rate chart shown in Figure~\ref{fig:lines} (right), including the blue and red background:
%
\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{brexitRate}~(\ident{lo}, \ident{hi})~\ident{rates}~=\kvd{overlay}~[\\
\quad \kvd{fill}~\strf{\#1F77B460}, [\;\,\kvd{cont}~\num{0}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{0}, \kvd{cont}~\ident{hi}\;],\\
\quad \kvd{fill}~\strf{\#D6272860}, [\;\,\kvd{cont}~\num{16}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{lo}, \kvd{cont}~\num{44}, \kvd{cont}~\ident{hi}, \kvd{cont}~\num{16}, \kvd{cont}~\ident{hi}\;],\\
\quad \kvd{line}~\strf{\#202020}~\ident{rates}~]\\[0.5em]
\kvd{overlay}~[~
\kvd{nest}_y~(\kvd{cont}~\num{0}),(\kvd{cont}~\num{50}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.25}, \num{1.50})~\ident{gbpusd}))))\\
\hspace{3.68em} \kvd{nest}_y~(\kvd{cont}~\num{50}),(\kvd{cont}~\num{100}), (\kvd{axis}_l~(\kvd{axis}_r~(\kvd{axis}_b~(\ident{brexitRate}~(\num{1.15}, \num{1.30})~\ident{gbpeur}))))~]\\
\end{array}
\end{equation*}

\noindent
Here, we use the \fkvd{let} binding of the host language to define a function that takes
a the data \ident{rates} together with the minimum and maximum. This is used for drawing
two filled rectangles, covering the first 16 days of the view in blue and the rest in red.
The shapes combined using \kvd{overlay} are rendered in the order in which they appear and so
the line shape is last, so that it appears above the background.

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{title}~t~s~=\ident{overlay}~[\\
\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
  (\kvd{nest}_y~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{15}), \\
\quad\quad \kvd{explicitScale}_x~(\kvd{continuous}~\num{0},\num{100}),~(\kvd{explicitScale}_y~(\kvd{continuous}~\num{0},\num{100}),\\
\quad\qquad \kvd{text}~\strf{\#000000}, (\kvd{cont}~\num{50}), (\kvd{cont}~\num{50}),~t)~)\\
\quad \kvd{nest}_x~(\kvd{cont}~\num{0}), (\kvd{cont}~\num{100}),
  (\kvd{nest}_y~(\kvd{cont}~\num{15}), (\kvd{cont}~\num{100}),~s)~]
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{l}
\fkvd{let}~\ident{pairplot}~\ident{attrs}~\ident{data}~=~\kvd{overlay}~[\\
\quad \kvd{for}~x~\kvd{in}~\ident{attrs}~\rightarrow~\kvd{for}~y~\kvd{in}~\ident{attrs}~\rightarrow\\
\qquad \kvd{nest}_x~(\kvd{cat}~x, 0), (\kvd{cat}~x, 1), (\kvd{nest}_y~(\kvd{cat}~y, 0), (\kvd{cat}~y, 1), \\
\quad\qquad (\kvd{axis}_l~(\kvd{axis}_b~(\kvd{overlay}~[~\fkvd{for}~v~\kvd{in}~\ident{data}~\rightarrow~\kvd{bubble}~(\ident{get}~x~v), (\ident{get}~y~v), \num{1}, \num{1} ])))~)]\\
\end{array}
\end{equation*}

let
  Shape.Layered [
    for x in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
      for y in ["sepal.length"; "sepal.width"; "petal.length"; "petal.width" ] do
        yield Shape.OuterScale(Some(Categorical [| ca x |]), Some(Categorical [| ca y |]),
          Shape.Axes(false, false, true, true, Shape.Layered [
            for v, k in iris ->
              Derived.StrokeColor(colors.[k], Shape.Bubble(numv v.[x], numv v.[y], 1., 1.))
          ]))
  ] |> render "out3"

@@

@@

@@

~
~
\noindent
[TODO: We really need illustrations for those example charts, but that would be more work!]
~
[TODO: Maybe use this chart as a motivation: https://wellcomeopenresearch.org/articles/4-63/v1]
~
\section{Random ideas}
\begin{itemize}
\item We could define a type system to catch categorical/continuous value mismatch in a single %scale.
\item It would be worth thinking about ordinal values, which are categorical but can be sorted.
\item There might be other kinds of scales - for example, color scale (can have meaning in scatter plot)
 or a scale for secondary markers (like sizes of bubbles in a bubble chart). We could really say that
 every value (including bar chart colors) should be coming from some scale...
\item Implementing something like \kvd{axis} or \kvd{roundScale} as an actual derived primitive is
 a bit tricky, because it needs to invoke a part of the normal rendering workflow (to run the
 automatic inference of scales on the nested shape) -- this might be just implementation issue,
 but it could be some more basic problem.
\end{itemize}
\end{document}
